Multi-level modeling with Openflexo/\FML

### 2022-02-23 8-discussion.tex
style guidelines: quotation marks -> \enquote
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

  Our solution uses the basic modeling constructs depicted by the \FML core
  metamodel (Figure~\ref{fig:mm}) and described in
  the section~\ref{sec:technology}. Models are \emph{virtual models}. They are
  composed of \emph{concepts}, being themselves concepts. Concepts may have
  \emph{roles} and \emph{behaviors} (actions one can perform).

  In order to provide the graphical tools, our solution also uses \emph{model
  slots} to connect some concept instances to external elements (in our case:
  instances of diagram from our diagraming TA). The use of slots has been
  described in the section~\ref{sec:model} and is illustrated by bold circles in
  the Figure~\ref{fig:ToolingArchitecture}.

@@

  The Openflexo approach is level agnostic: ``levels'' have no specific nature
  and there are no numbered levels. In our solution, concepts of a given level
  are grouped into a virtual model. Inheritance and instantiation allow the
  establishment of relationships between concepts from different levels.
%\ref{fig:BPMNSubsetExample}
%  Figure~\ref{fig:AcmeFullArchitecture} shows three levels: the metamodel, the
%  ACME metamodel and the ACME model.
Figure~\ref{fig:MultilevelArchitecture} shows the multi-level architecture of
our solution.

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.
  For instance, in the figure~\ref{fig:AcmeFullArchitecture}, the
  \texttt{Developer} concept from \texttt{Acme} metamodel specializes the
  \texttt{ActorType} of the base metamodel. The \texttt{Software Engineering Process} of the
  \texttt{Acme} model is an instance of \texttt{ProcessType} in the metamodel.
%  \textcolor{red}{Phrase faite a relire sur Developer\dots}

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models. Section~\ref{sec:model} describes
  how the requirements are captured and how behaviors are implemented with
  \FML. For instance, \texttt{isAuthorizedActor} behavior is needed to enforce
  \textbf{P17} requirement, leading to writing \FML code in \texttt{TaskType}
  and \texttt{Task}. Note that in our approach, metamodels are built in an
 ad hoc way together with the models (co-construction), therefore the user
  also validates on an on-demand basis.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts. This
  type of change can be done without much effort. On the other hand, making
  changes to the existing models could be difficult. For example, given a
  requirement, an instance cannot be changed into a concept easily. This case
  could occur if the challenge had refined the specification by requiring
  specific design tasks.  In this context, we would have to transform the
  \inst{Design Task} instance into a concept whose instances could have been
  "using agile methodology".

>>>>

  Our solution uses the basic modeling constructs depicted by the \FML core
  metamodel (Fig.~\ref{fig:mm}) and described in
  the Sect.~\ref{sec:technology}. Models are \emph{virtual models}. They are
  composed of \emph{concepts}, being themselves concepts. Concepts may have
  \emph{roles} and \emph{behaviors} (actions one can perform).

  In order to provide the graphical tools, our solution also uses \emph{model
  slots} to connect some concept instances to external elements (in our case:
  instances of diagram from our diagraming TA). The use of slots has been
  described in the Sect.~\ref{sec:model} and is illustrated by bold circles in
  the Fig.~\ref{fig:ToolingArchitecture}.

@@

  The Openflexo approach is level agnostic: ``levels'' have no specific nature
  and there are no numbered levels. In our solution, concepts of a given level
  are grouped into a virtual model. Inheritance and instantiation allow the
  establishment of relationships between concepts from different levels.
%\ref{fig:BPMNSubsetExample}
%  Fig.~\ref{fig:AcmeFullArchitecture} shows three levels: the metamodel, the
%  ACME metamodel and the ACME model.
Fig.~\ref{fig:MultilevelArchitecture} shows the multi-level architecture of
our solution.

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.
  For instance, in the Fig.~\ref{fig:AcmeFullArchitecture}, the
  \texttt{Developer} concept from \texttt{Acme} metamodel specializes the
  \texttt{ActorType} of the base metamodel. The \texttt{Software Engineering Process} of the
  \texttt{Acme} model is an instance of \texttt{ProcessType} in the metamodel.
%  \textcolor{red}{Phrase faite a relire sur Developer\dots}

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models. Sect.~\ref{sec:model} describes
  how the requirements are captured and how behaviors are implemented with
  \FML. For instance, \texttt{isAuthorizedActor} behavior is needed to enforce
  \textbf{P17} requirement, leading to writing \FML code in \texttt{TaskType}
  and \texttt{Task}. Note that in our approach, metamodels are built in an
 ad hoc way together with the models (co-construction), therefore the user
  also validates on an on-demand basis.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts. This
  type of change can be done without much effort. On the other hand, making
  changes to the existing models could be difficult. For example, given a
  requirement, an instance cannot be changed into a concept easily. This case
  could occur if the challenge had refined the specification by requiring
  specific design tasks.  In this context, we would have to transform the
  \inst{Design Task} instance into a concept whose instances could have been
  \enquote{using agile methodology}.

### 2022-02-23 9-relatedworks.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

A first description of process modeling as a multi-level modeling problem was proposed by~\citet{lara2018refactoring} in the context of a catalogue of refactoring for multi-level models (in a simpler form with fewer constraints and requirements w.r.t. the challenge version). A solution is provided with MetaDepth~\parencite{metadepth}, which supports modeling with any number of levels, dual ontological/linguistic typing and deep characterization through potency~\citep{potency}. For their solution to the multi-level process problem they use three levels. The first level describes generic processes, the second level software engineering processes, and the third level, software engineering enacted processes. Additionally, the authors use~\emph{linguistic extensions}~\citep{metadepth}, a mechanism to  linguistically extend ontological instance models at any level, in order to introduce artifact types and tasks duration in the second level. Our solution is similar to theirs w.r.t. the grouping and organization of concepts (e.g., a model for general process, concepts, an extension for software engineering process) but it replaces the use of clabjets and potency with the use of the type-object pattern~\citep{typeObject}. The so-called \emph{linguistic extensions} are natively supported in Openflexo/\FML.

@@

More similar to us,~\citet{deeptelos2019} uses DeepTelos \parencite{deeptelos2016}, an extension of the Telos language \parencite{telos1990} in order to solve the process challenge. As with Openflexo/\FML, they do not use explicit (numbered) levels nor potency. However, unlike Openflexo, DeepTelos integrates a multi-level modeling specific construct similar to the PowerType \parencite{atkinson2001essence} pattern they call \emph{most general instances} which the authors use to simulate potency. Models in DeepTelos are organized in (tree-like) module hierarchies, where submodules can see all the concepts defined in parent modules. The authors use this module system in order to \emph{organize} their solution for the Multi-level process challenge. Concretely they created a hierarchy of modules such as the top module contains base concepts and formulas required for multi-level modeling (e.g., the support for \emph{most general instances}), and a sub-module contains process definitions fulfilling requirements P1 to P19. This sub-module contains in turn a sub-module representing the coding process type whereas concrete coding processes are represented in subsequent sub-modules. There are two main differences between the DeepTelos and Openflexo/\FML: 1) we use the type-object pattern instead of the powertype pattern; 2) DeepTelos reifies their support for multi-level modeling while our remains ad hoc.

>>>>

A first description of process modeling as a multi-level modeling problem was
proposed by~\citet{lara2018refactoring} in the context of a catalogue of
refactoring for multi-level models (in a simpler form with fewer constraints
and requirements w.r.t. the challenge version). A solution is provided with
MetaDepth~\parencite{metadepth}, which supports modeling with any number of
levels, dual ontological/linguistic typing and deep characterization through
potency~\citep{potency}. For their solution to the multi-level process problem
they use three levels. The first level describes generic processes, the second
level software engineering processes, and the third level, software engineering
enacted processes. Additionally, the authors use~\emph{linguistic
extensions}~\citep{metadepth}, a mechanism to  linguistically extend
ontological instance models at any level, in order to introduce artifact types
and tasks duration in the second level. Our solution is similar to theirs
w.r.t. the grouping and organization of concepts (\eg, a model for general process, concepts, an extension for software engineering process) but it replaces the use of clabjets and potency with the use of the type-object pattern~\citep{typeObject}. The so-called \emph{linguistic extensions} are natively supported in Openflexo/\FML.

@@

More similar to us,~\citet{deeptelos2019} uses DeepTelos
\parencite{deeptelos2016}, an extension of the Telos language
\parencite{telos1990} in order to solve the process challenge. As with
Openflexo/\FML, they do not use explicit (numbered) levels nor potency.
However, unlike Openflexo, DeepTelos integrates a multi-level modeling specific
construct similar to the PowerType \parencite{atkinson2001essence} pattern they
call \emph{most general instances} which the authors use to simulate potency.
Models in DeepTelos are organized in (tree-like) module hierarchies, where
submodules can see all the concepts defined in parent modules. The authors use
this module system in order to \emph{organize} their solution for the
Multi-level process challenge. Concretely they created a hierarchy of modules
such as the top module contains base concepts and formulas required for
multi-level modeling (\eg, the support for \emph{most general instances}), and a sub-module contains process definitions fulfilling requirements P1 to P19. This sub-module contains in turn a sub-module representing the coding process type whereas concrete coding processes are represented in subsequent sub-modules. There are two main differences between the DeepTelos and Openflexo/\FML: 1) we use the type-object pattern instead of the powertype pattern; 2) DeepTelos reifies their support for multi-level modeling while our remains ad hoc.
